Generalized Linear Mixed Effects Models

### 2024-03-02 posts/_12_Examples_LatentTransitionAnalysis.qmd
img uploads

---
title: "Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables

@@

------

**STEP 1: Convert dataframe from long-to-wide format**
```{r long-to-wide}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Split the data into static (unchanging) and changing variables
# Static variables are taken only from the baseline measurement occasion
df_static <- df %>%
  filter(eventname == "baseline_year_1_arm_1") %>%
  select(src_subject_id, site_id_l, sex)

df_changing <- df %>%
  select(src_subject_id, eventname, anthroweightcalc, anthroheightcalc)

# Convert the changing variables from long format to wide format
df_wide_changing <- df_changing %>%
  gather(variable, value, -src_subject_id, -eventname) %>%
  unite("variable_event", variable, eventname) %>%
  spread(variable_event, value)

# Merge the static and changing data frames based on the 'src_subject_id' variable
df_wide <- df_static %>%
  left_join(df_wide_changing, by = "src_subject_id")

# Rename columns to shorter names for easier reference
df_wide <- df_wide %>% 
        rename("anthroweightcalc_0"="anthroweightcalc_baseline_year_1_arm_1",
               "anthroweightcalc_1"="anthroweightcalc_1_year_follow_up_y_arm_1",
               "anthroweightcalc_2"="anthroweightcalc_2_year_follow_up_y_arm_1",
               "anthroweightcalc_3"="anthroweightcalc_3_year_follow_up_y_arm_1",
               "anthroweightcalc_4"="anthroweightcalc_4_year_follow_up_y_arm_1",
               "anthroheightcalc_0"="anthroheightcalc_baseline_year_1_arm_1",
               "anthroheightcalc_1"="anthroheightcalc_1_year_follow_up_y_arm_1",
               "anthroheightcalc_2"="anthroheightcalc_2_year_follow_up_y_arm_1",
               "anthroheightcalc_3"="anthroheightcalc_3_year_follow_up_y_arm_1",
               "anthroheightcalc_4"="anthroheightcalc_4_year_follow_up_y_arm_1"
               )

# Reorder columns in the dataframe
df_wide <- df_wide %>% select(src_subject_id, site_id_l, sex, anthroweightcalc_0, anthroweightcalc_1, anthroweightcalc_2, anthroweightcalc_3, anthroweightcalc_4, anthroheightcalc_0, anthroheightcalc_1, anthroheightcalc_2, anthroheightcalc_3, anthroheightcalc_4)

@@

The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname'). It incorporates xxxxxx. The results of the model are then printed to provide a detailed summary of the fitted model parameters.

**STEP 2: Compute xxxxx**

```{r}

# Define the Latent Growth Curve Model
model <- '
  # Latent variables
    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2

  # Estimate means of latent variables
    i ~ 1
    s ~ 1

  # Variances
    i ~~ i
    s ~~ s

  # Covariance
    i ~~ s
'
```

```{r}
# Fit the model
fit <- sem(model, data = df_wide, missing = "FIML")

>>>>

---
title: "Latent Transition Analysis"
author: "Biostatistics Working Group"
---

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT", "tidyLPA")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(tidyLPA)      # XXXXXXXX

@@

**STEP 1: XXXXXXXXX **

-------

```{r}
# Estimate the LTA model
fit <- tidyLPA::estimate_profiles(data, 2, variances = "equal", covariances = "zero")

plot(fit)

@@

### 2024-03-02 posts/_14_Examples_StateTraitModels.qmd
img uploads

---
title: "Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

@@

```{r}

# Define the Latent Growth Curve Model
model <- '
  # Latent variables
    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2

  # Estimate means of latent variables
    i ~ 1
    s ~ 1

  # Variances
    i ~~ i
    s ~~ s

  # Covariance
    i ~~ s
'
```

```{r}
# Fit the model
fit <- sem(model, data = df_wide, missing = "FIML")

>>>>

---
title: "State-Trait Models"
author: "Biostatistics Working Group"
---

@@

```{r}
# Define the model
model <- '
  # Trait component
  trait =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3

  # State component
  state =~ anthroheightcalc_0 + anthroheightcalc_1 + anthroheightcalc_2 + anthroheightcalc_3

  # Residual variances
  anthroheightcalc_0 ~~ anthroheightcalc_0
  anthroheightcalc_1 ~~ anthroheightcalc_1
  anthroheightcalc_2 ~~ anthroheightcalc_2
  anthroheightcalc_3 ~~ anthroheightcalc_3
'
```

```{r}
# Estimate the State-Trait Model
fit <- sem(model, data = df_wide, missing = "FIML")

### 2024-03-02 posts/_6_Examples_GeneralizedEstimatingEquations.qmd
img uploads

---
title: "State-Trait Models"
author: "Biostatistics Working Group"
---

# XXXXXXXXXXX {.unnumbered}

@@

XXXXXXXXXXX Models xxxxxxx. This xxxxxxx.

In this example, we will utilize xxxxxx models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time. XXXX models facilitate this by xxxxx.

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables

@@

------

**STEP 1: Convert dataframe from long-to-wide format**
```{r long-to-wide}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Split the data into static (unchanging) and changing variables
# Static variables are taken only from the baseline measurement occasion
df_static <- df %>%
  filter(eventname == "baseline_year_1_arm_1") %>%
  select(src_subject_id, site_id_l, sex)

df_changing <- df %>%
  select(src_subject_id, eventname, anthroweightcalc, anthroheightcalc)

# Convert the changing variables from long format to wide format
df_wide_changing <- df_changing %>%
  gather(variable, value, -src_subject_id, -eventname) %>%
  unite("variable_event", variable, eventname) %>%
  spread(variable_event, value)

# Merge the static and changing data frames based on the 'src_subject_id' variable
df_wide <- df_static %>%
  left_join(df_wide_changing, by = "src_subject_id")

# Rename columns to shorter names for easier reference
df_wide <- df_wide %>% 
        rename("anthroweightcalc_0"="anthroweightcalc_baseline_year_1_arm_1",
               "anthroweightcalc_1"="anthroweightcalc_1_year_follow_up_y_arm_1",
               "anthroweightcalc_2"="anthroweightcalc_2_year_follow_up_y_arm_1",
               "anthroweightcalc_3"="anthroweightcalc_3_year_follow_up_y_arm_1",
               "anthroweightcalc_4"="anthroweightcalc_4_year_follow_up_y_arm_1",
               "anthroheightcalc_0"="anthroheightcalc_baseline_year_1_arm_1",
               "anthroheightcalc_1"="anthroheightcalc_1_year_follow_up_y_arm_1",
               "anthroheightcalc_2"="anthroheightcalc_2_year_follow_up_y_arm_1",
               "anthroheightcalc_3"="anthroheightcalc_3_year_follow_up_y_arm_1",
               "anthroheightcalc_4"="anthroheightcalc_4_year_follow_up_y_arm_1"
               )

# Reorder columns in the dataframe
df_wide <- df_wide %>% select(src_subject_id, site_id_l, sex, anthroweightcalc_0, anthroweightcalc_1, anthroweightcalc_2, anthroweightcalc_3, anthroweightcalc_4, anthroheightcalc_0, anthroheightcalc_1, anthroheightcalc_2, anthroheightcalc_3, anthroheightcalc_4)

@@

**STEP 2: Compute xxxxx**

```{r}
# Define the model
model <- '
  # Trait component
  trait =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3

  # State component
  state =~ anthroheightcalc_0 + anthroheightcalc_1 + anthroheightcalc_2 + anthroheightcalc_3

  # Residual variances
  anthroheightcalc_0 ~~ anthroheightcalc_0
  anthroheightcalc_1 ~~ anthroheightcalc_1
  anthroheightcalc_2 ~~ anthroheightcalc_2
  anthroheightcalc_3 ~~ anthroheightcalc_3
'
```

```{r}
# Estimate the State-Trait Model
fit <- sem(model, data = df_wide, missing = "FIML")

summary(fit, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)

>>>>

---
title: "Generalized Estimating Equations"
author: "Biostatistics Working Group"
---

# Generalized Estimating Equations {.unnumbered}

@@

This example assesses xxxxxxxx.

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT", "geepack")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(geepack)      # Generalized Estimating Equations (GEE) analysis
```
:::

@@

@@

**STEP 1: Compute xxxxx**
```{r}
model <- geeglm(anthroweightcalc ~ eventname, data = df, id = src_subject_id, family = gaussian(), corstr = "exchangeable")
```

```{r}
summary(model)
```

### 2024-03-02 posts/_7_Examples_GeneralizedLinearMixedEffectsModels.qmd
img uploads

---
title: "Generalized Estimating Equations"
author: "Biostatistics Working Group"
---

# Generalized Estimating Equations {.unnumbered}

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT", "geepack")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(geepack)      # Generalized Estimating Equations (GEE) analysis
```
:::

@@

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
subset_data <- data_nonimaging[, selected_vars]

@@

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

@@

# Define a function to compute descriptives
compute_descriptives <- function(data, event_name) {
  # For factor variables
  sex_desc <- paste0(table(data$sex), " (", round(100 * prop.table(table(data$sex)), 1), "%)")
  race_desc <- paste0(table(data$race.4level), " (", round(100 * prop.table(table(data$race.4level)), 1), "%)")

  # For numeric variables
  age_desc <- paste0(round(mean(data$age, na.rm = TRUE), 2), " (", round(sd(data$age, na.rm = TRUE), 2), ")")
  weight_desc <- paste0(round(mean(data$anthroweightcalc, na.rm = TRUE), 2), " (", round(sd(data$anthroweightcalc, na.rm = TRUE), 2), ")")
  height_desc <- paste0(round(mean(data$anthroheightcalc, na.rm = TRUE), 2), " (", round(sd(data$anthroheightcalc, na.rm = TRUE), 2), ")")

  # Combine into a data frame
  desc_df <- data.frame(
    Variable = c("Sex - Female", "Sex - Male or other", "Race - Asian", "Race - Black", "Race - Other/Mixed", "Race - White", "Age", "Weight", "Height"),
    Value = c(sex_desc, race_desc, age_desc, weight_desc, height_desc)
  )

@@

**STEP 1: Compute xxxxx**
```{r}
model <- geeglm(anthroweightcalc ~ eventname, data = df, id = src_subject_id, family = gaussian(), corstr = "exchangeable")
```

>>>>

---
title: "Generalized Linear Mixed Effects Models"
author: "Biostatistics Working Group"
---

# Generalized Linear Mixed Effects Models {.unnumbered}

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT", "lme4")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(lme4)         # Linear mixed-effects models
```
:::

@@

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "tlfb_alc_sip_l")
subset_data <- data_nonimaging[, selected_vars]

@@

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    tlfb_alc_sip_l = as.factor(tlfb_alc_sip_l)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

@@

# Define a function to compute descriptives
compute_descriptives <- function(data, event_name) {
  # For factor variables
  sex_desc <- paste0(table(data$sex), " (", round(100 * prop.table(table(data$sex)), 1), "%)")
  race_desc <- paste0(table(data$race.4level), " (", round(100 * prop.table(table(data$race.4level)), 1), "%)")
  tlfb_alc_sip_l_desc <- paste0(table(data$tlfb_alc_sip_l), " (", round(100 * prop.table(table(data$tlfb_alc_sip_l)), 1), "%)")

  # For numeric variables
  age_desc <- paste0(round(mean(data$age, na.rm = TRUE), 2), " (", round(sd(data$age, na.rm = TRUE), 2), ")")
  weight_desc <- paste0(round(mean(data$anthroweightcalc, na.rm = TRUE), 2), " (", round(sd(data$anthroweightcalc, na.rm = TRUE), 2), ")")

  # Combine into a data frame
  desc_df <- data.frame(
    Variable = c("Sex - Female", "Sex - Male or other", "Race - Asian", "Race - Black", "Race - Other/Mixed", "Race - White", "tlfb_alc_sip_l - No", "tlfb_alc_sip_l - Yes", "Age", "Weight"),
    Value = c(sex_desc, tlfb_alc_sip_l_desc, race_desc, age_desc, weight_desc)
  )

@@

**STEP 1: Compute xxxxx**
```{r}
model <- glmer(tlfb_alc_sip_l ~ eventname + (1 | src_subject_id), data = df, family = binomial(link = "logit"))
```
